---
title: "Data 607 Homework 6 by Jeff Nieman"
date: "March 6, 2016"
output: html_document
---

The process of creating tidy began with uploading the .csv file I created onto github and loading the various packages required.  I added na.strings="" to place NA in the fields that would have been blank.

```{r}
flightdata <- read.csv('https://raw.githubusercontent.com/jeffnieman11/Data607HW6/master/flight_delay.csv',na.strings="")
flightdata
require(tidyr)
require(zoo)
require(dplyr)
require(ggplot2)
```

Using the gather function I created a column for the key "city" and another for the value "n".  I then renamed the column headers to better reflect what the data show.

```{r}
flightdata1 <- gather(flightdata, "city", "n", 3:7)
colnames(flightdata1) <- c("Airline", "Timeliness", "City", "Count")
flightdata1
```

Using the na.locf function from the zoo package to fill the NA fields with the value above it.

```{r}

flightdata2 <- na.locf(flightdata1)
flightdata2

```

Using the spread function I separated out the timeliness column into new columns separating the ontime and delayed values.  I then renamed the columns and made sure the type for the delayed and ontime columns of data was numeric.

```{r}
flightdata3 <- spread(flightdata2, Timeliness, Count)
colnames(flightdata3) <- c("Airline", "City", "Delayed", "Ontime")
flightdata3$Delayed <- as.numeric(as.character(flightdata3$Delayed))
flightdata3$Ontime <- as.numeric(as.character(flightdata3$Ontime))
flightdata3
```

Using the mutate function I created a new column called "Rate".  This is the percentage of delayed flights for each airline by each run.

```{r}
flightdata4 <- mutate(flightdata3, Rate = Delayed/(Delayed + Ontime))
flightdata4 <- arrange(flightdata4, desc(Rate))
flightdata4
```

Using ggplot2 I created a bar chart that shows the delayed rate for each city by airline.  It shows clearly that Amwest was more often delayed than Alaska in every city.

```{r}
ggplot(flightdata4,aes(x=City, y=Rate, fill=factor(Airline))) +
  geom_bar(stat="identity",position="dodge") +
  scale_fill_discrete(name="Airline", breaks =c("ALASKA", "AMWEST"), labels=c("Alaska", "Amwest")) +
  xlab("City") +
  ylab("Delayed Rate")
```

Using the group_by and summarise functions I created a summary for all cities by airline.  Then I created the delayed rate for each airline using the mutate function.

```{r}
flightdata5 <- flightdata4 %>%
  group_by(Airline)  %>%
  summarise(Delayed = sum(Delayed), Ontime = sum(Ontime))
flightdata5

flightdata6 <- mutate(flightdata5, Rate = Delayed/(Delayed + Ontime))
flightdata6 
```

Finally once again using ggplot2 I created a bar chart showing the delay rate for all cities for each of the two airlines.  Again this clearly shows that Amwest overall was delayed more often than Alaska.

```{r}
ggplot(flightdata6,aes(x=Airline, y=Rate)) +
  geom_bar(stat="identity",position="dodge") +
  xlab("Airline") +
  ylab("Delayed Rate")
```

In conclusion, in every city and overall Amwest has a higher percentage of delayed flights than Alaska.

